From Real-time Attention Assessment to ``With-me-ness'' in Human-Robot Interaction

### 2015-09-30 paper.tex
Languages fix from James -- Part 2

To assess the accuracy of the attention estimation by the robot, we computed
over time (excluding the periods when the head is not detected) the
overlap between the ground-truth and the robot's estimate and the
inter-rater agreement (Cohen's $\kappa$).
%and the cosine similarity
%matrix.

@@

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=\linewidth]{matches-excerpt}
    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
        truth} during a face-to-face child-robot interaction (subject 4
        in table~\ref{tab:results}, 3.5min-long excerpt).
        In blue (top lines), the focus of attention as computed by the robot;
        in orange (bottom lines), the focus of attention as manually annotated
        (ground-truth). Periods where the head is lost are marked with dashed
        vertical lines. The bottom graph shows agreement between both (whenever
        the head is detected).}

@@

During the whole interaction, the head pose of the children was satisfactorily
and consistently tracked (86\% of the time in average, $SD=3.0$). While this
high score is expected for a face-to-face interaction with a static head-tracking camera
(meaning that the child head would remain in the field of view of the camera
most of the time), this is still comforting in terms of suitability of our
approach for head pose estimation with children in field experiments of this
kind. Expectedly, the primary causes of lost head pose were occlusions with the
hands (similar to the middle-bottom picture in
Figure~\ref{head_pose_real_world}), close proximity with the tablet while
writing, and gaze directed to the facilitator (who was sitting directly on the left of
the child, Figure~\ref{fig:realSetup}).

In terms of attention tracking, Cohen's $\kappa$ values are between 0.47 and
0.68 with one subject resulting in significantly worst tracking, at 0.26. While
the interpretation of Cohen's $\kappa$ is subject to discussions (the number of
the coded values -- in our case 6 -- and the distribution probability of values
-- in our case, values are \emph{not} equiprobable -- are factors impacting
$\kappa$ independently of the level of agreement), the levels of agreement are
\emph{moderate} to \emph{substantial}, with one subject only
showing \emph{fair} agreement~\cite{landis1977measurement}. Further analysis of
the videos shows that the child with the lowest level of agreement is especially
calm and would rely more on the eyes to direct his gaze than the other children,
thus leading to a less accurate estimation of its focus of attention.

@@

Over the course of the experiment presented at section~\ref{sec:expe}, the robot
controller would associate a set of expected attentional targets to the phase of
the interaction (Table~\ref{tab:targets}). For instance, while the robot was
waiting for the child's handwriting demonstration (``\textit{Waiting for
feedback}''), the expected attentional target of the child was the tablet (since
the child was supposed to write there) or the secondary tablet (that displayed a
template of the word, used as reference by the child). These expected targets
form altogether the robot's attentional {\it a priori} and are used to compute
the with-me-ness.  They are represented over time by the green lines on
Figure~\ref{fig:with-me-ness}. With-me-ness can be calculated over the whole
interaction (see corresponding results in Table~\ref{tab:results-with-me-ness})
or over shorter time windows. Shorter time windows are interesting for two
purposes: to analyse the level of with-me-ness in relation to specific
interaction episodes; to allow a measurement of with-me-ness by the robot
\emph{over the course} of the interaction (\emph{in-the-moment} measurement) --
in this later case, one may typically want to consider a sliding time window.

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{with-me-ness}
    \caption{\small \textbf{With-me-ness}. Evolution of the level of
        \emph{with-me-ness} over the whole $\approx$17min long interaction of
        subject 2. The top chart is similar to Figure~\ref{fig:realExpected}
        with the \emph{expected} attentional targets added in green. The bottom
        diagram represents the instantaneous level of with-me-ness over a
        sliding window of 30 seconds. The blue line is with-me-ness as estimated
        by the robot, the orange line is with-me-ness computed from manually
        annotated attentional targets.  Pearson's correlation between both: $r(973)=0.58, p =
        2.2e^{-16}.$}

@@

\begin{table}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness
    level is reported over the whole interaction, either based on the
    ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
    attention measured by the robot. Pearson's correlation: $r(4)=0.46$.}

@@

A chart like Figure~\ref{fig:with-me-ness} remains an useful tool to analyse the
interaction, and several observations can be made out of it: as said, the green
lines represent how the robot imagine, at a given time, the attention
distribution of the child. They also provide an accurate picture of the overall
turn-taking as viewed by the robot: for instance, the episode ``B'' on
Figure~\ref{fig:with-me-ness} corresponds to one of the ``\emph{Robot writing}''
episodes, surrounded by ``\emph{Waiting for feedback}'' phases like ``C''; episode
``D'' corresponds to the story telling; etc. In terms of interaction, the large
variance of the duration of these phases reflects the fact that this child would
sometimes take a lot of time to send feedback to the robot, and sometimes, on
the contrary, be very quick.

Looking at the ground-truth focus of attention (orange lines), the first
striking observation is that this child did generally \emph{closely} follow what
the robot was expecting: in that regard, it seems that the child was very much
engaged into the interaction (we discuss in the next section the exact
relationships between with-me-ness and engagement). The only major exception is
the story-telling phase (episode ``D''): the child was seemingly not captivated by the
first half of the story, and his attention was not directed towards the robot.

Another interesting observation pertains to the facilitator: as one can see,
this child only rarely turned to the facilitator, indicating that the
interaction and the task were meaningful and easy for him to follow alone.

More subtle patterns and events can also be observed: for instance, during the
feedback phases like episode ``C'', we
can notice numerous gaze shifts between the tablet (where the child write) and the
secondary tablet (that showed a template of the word). The episode ``B'' (robot
writing) is also interesting: the child did not look at the robot, and instead
remained focused on the secondary tablet. This situation is typically useful for
the robot to detect as it may want to adapt its behavior to recover the child's
attention.

@@

As hinted by the title of this article, we have presented here how a robot can
effectively assess in real-time and with a regular camera the focus of attention
of its interactant, and how we can, in turn, combine it with the robot's {\it a priori}
about the interaction to build a metric of
\emph{with-me-ness} over the course of the interaction.

@@

We already stated the main limitations of our approach to estimation of the
focus of attention: eye gaze information is neglected and we do not perform
actual visibility check of the in focus objects (we simply approximate them to
their origins, ignoring possible occlusions).

While the first issue is shared with most of the other vision (2D or 3D) or motion
capture techniques for real-time gaze estimation found in robotics,
our results are positive: we shows that relying
purely on head pose estimation to estimate gaze direction leads to real-world
measures that are worth being considered and used. They may not match manual
annotations, but they are definitely a valuable \emph{in-the-moment} input for
the robot, reaching for certain children levels of accuracy traditionally
considered as very good.

@@

Our approach relies on a simple, non-intrusive sensor (a RGB camera on
the robot) and an open-source, fast pose estimation algorithm\fixme{test on nao!
how fast is it actually?}: we hope that this may contribute to a widespread
adoption of such a technique on a range of robots, including the common {\sc
nao} platform.

@@

Borrowing the neologism from the field of CSCL, we have also introduced in this
article
\emph{with-me-ness} as a measure of ``how much the user in \emph{with} the robot during
a task''. This can be acquired over the course the interaction,
thus providing the robot with a real-time metric for relatively high-level social construct,
undoubtedly related to engagement.

One may reasonably wonder how different is \emph{with-me-ness} from \emph{joint
attention} on one hand, and from \emph{engagement} on the other hand.
With-me-ness is related to both, with however noteworthy nuances: (Triadic) \emph{joint
attention} is understood as the cognitive realization of a shared attention to
an object, itself building on a shared perception of that object (\ie joint attention
builds on a \emph{perceptual} alignment of two agents). \emph{Conceptual
with-me-ness} as proposed by Sharma~\etal is on the contrary \emph{referential}: ``you are
\emph{with} me if you focus on what I refer to, either explicitly or
implicitly''. We propose to understand it here in an even slightly broader sense, that
better reflect the \emph{interaction}: ``you are \emph{with} me if you focus on
what is important for the interactive task at hand.''

On the other hand, with-me-ness is only a precursor of engagement: it does not
say much about the \emph{cognitive} commitment of a user to an interaction. A
user may closely adhere to the injunctions of the robot (or, actually, of the
experimenters), with thus high levels of with-me-ness,  without being
\emph{engaged} into the interaction. This is typically seen in child-robot
interaction: children will attempt to closely follow what they are asked to do
-- which may \emph{look like} they are engaged into the interaction -- while
they merely \emph{obey orders}.

Compared to engagement, one of the strength of with-me-ness is its specificity:
it is well-defined (the only important variable -- that needs to be reported --
is the mapping {\it phase/expected attentional target}, \ie our
Table~\ref{tab:targets}), we can formalize it, and as such, it is valuable to
assess and compare how users are willing or able to interact with a robot.
Besides, we have hopefully demonstrated in this article that with-me-ness is an
operational \emph{in-the-moment} metric that can be also used as feedback to the
robot controller to build richer, more adaptive interactive behaviors for our
robots.

>>>>

To assess the accuracy of the attention estimation by the robot, we computed
over time the overlap between the ground-truth and the robot's estimate and the
inter-rater agreement (Cohen's $\kappa$). The periods where the head was not
detected were \emph{excluded} from the agreement computation: at such times, the
robot explicitly knows that it can not estimate the focus of attention, and as
such, we do not consider that it \emph{wrongly} estimates the focus.

%and the cosine similarity
%matrix.

@@

\begin{figure*}[ht!]
    \centering
    \includegraphics[width=\linewidth]{matches-excerpt}
    \caption{\small \textbf{Comparison of computed focus of attention \vs ground
        truth} during a face-to-face child-robot interaction (subject 4
        in table~\ref{tab:results}, 3.5min-long excerpt).
        In blue (top lines), the focus of attention as computed by the robot;
        in orange (bottom lines), the focus of attention as manually annotated
        (ground-truth). The bottom section shows agreement between both (whenever
        the head is detected).}

@@

During the whole interaction, the head pose of the children was consistently
tracked, 86\% of the time in average, $SD=3.0$. While this high score is
expected for a face-to-face interaction with a static head-tracking camera
(meaning that the child head would remain in the field of view of the camera
most of the time), this is still comforting in terms of suitability of our
approach for head pose estimation with children in field experiments of this
kind. Expectedly, the primary causes of lost head pose were occlusions with the
hands (similar to the middle-bottom picture in
Figure~\ref{head_pose_real_world}), close proximity with the tablet while
writing, and gaze directed to the facilitator (who was sitting directly on the
left of the child, Figure~\ref{fig:realSetup}).

In terms of attention tracking, Cohen's $\kappa$ values are between 0.47 and
0.68 with one subject resulting in significantly worst tracking, at 0.26. While
the interpretation of Cohen's $\kappa$ is subject to discussion (the number of
the coded values -- in our case 6 -- and the distribution probability of values
-- in our case, values are \emph{not} equiprobable -- are factors impacting
$\kappa$ independently of the level of agreement), the levels of agreement are
\emph{moderate} to \emph{substantial}, with one subject only showing \emph{fair}
agreement~\cite{landis1977measurement}. Further analysis of the videos shows
that the child with the lowest level of agreement was particularly quiet and
would indeed rely more on the eyes to direct his gaze than the other children,
thus leading to a less accurate estimation of his focus of attention.

@@

Over the course of the experiment presented in section~\ref{sec:expe}, the robot
controller would associate a set of expected attentional targets to the phase of
the interaction (Table~\ref{tab:targets}). For instance, while the robot was
waiting for the child's handwriting demonstration (``\textit{Waiting for
feedback}''), the expected attentional target of the child was the tablet (since
the child was supposed to write there) or the secondary tablet (that displayed a
template of the word, used as a reference by the child). These expected targets (green lines on Figure~\ref{fig:with-me-ness})
form the robot's attentional {\it a priori} knowledge and are used to compute
the with-me-ness. With-me-ness can be calculated over the whole
interaction
or over shorter time windows. With-me-ness over the whole interaction for the
six subjects is
reported in Table~\ref{tab:results-with-me-ness}. The Pearson's correlation with the
ground-truth is $r(4)=0.46$ (significance not computed due to small
sample size). Shorter time windows are interesting for two
purposes: to analyse the level of with-me-ness in relation to specific
interaction episodes; to allow a measurement of with-me-ness by the robot
\emph{over the course} of the interaction (\emph{in-the-moment} measurement) --
in the latter case, one may typically want to consider a sliding time window.

\begin{figure*}
    \centering
    \includegraphics[width=\linewidth]{with-me-ness}
    \caption{\small \textbf{With-me-ness}. Evolution of the level of
        \emph{with-me-ness} over the whole $\approx$17min long interaction of
        subject 2. The top chart is similar to Figure~\ref{fig:realExpected}
        with the \emph{expected} attentional targets added in green. The bottom
        diagram represents the instantaneous level of with-me-ness over a
        sliding window of 30 seconds. The blue line is the with-me-ness as estimated
        by the robot, the orange line is the with-me-ness computed from manually
        annotated attentional targets.  Pearson's correlation between both
        series for this subject: $r(973)=0.58, p < 0.001$.}

@@

\begin{table}[h!]
    \centering
    \caption{\textbf{Levels of with-me-ness}. For each subject, the with-me-ness
    level is reported over the whole interaction, either based on the
    ground-truth (\ie \emph{ground-truth with-me-ness}), or based on the focus of
    attention measured by the robot.}

@@

A chart like Figure~\ref{fig:with-me-ness} remains a useful tool to analyse the
interaction, and several observations can be made from it: the green
lines represent how the robot imagine, at a given time, the attention
distribution of the child. They also provide an accurate picture of the overall
turn-taking as viewed by the robot: for instance, the episode ``B'' on
Figure~\ref{fig:with-me-ness} corresponds to one of the ``\emph{Robot writing}''
episodes, surrounded by ``\emph{Waiting for feedback}'' phases like ``C''; episode
``D'' corresponds to the story telling; etc. In terms of interaction, the large
variance of the duration of these phases reflects the fact that this child would
sometimes take a lot of time to send feedback to the robot, and sometimes, on
the contrary, be very quick.

Looking at the ground-truth focus of attention (orange lines), the first
striking observation is that this child did generally \emph{closely} follow what
the robot was expecting: in that regard, it seems that the child was very much
engaged in the interaction (we discuss in the next section the exact
relationships between with-me-ness and engagement). The only major exception is
the story-telling phase (episode ``D''): the child was seemingly not captivated by the
first half of the story, and their attention was not directed towards the robot.

Another interesting observation pertains to the facilitator: as one can see,
this child only rarely turned to the facilitator, possibly indicating that the
interaction and the task were meaningful and easy enough for him to follow alone.

More subtle patterns and events can also be observed: for instance, during the
feedback phases like episode ``C'', we
can notice numerous gaze shifts between the tablet (where the child writes) and the
secondary tablet (that showed a template of the word). The episode ``B'' (robot
writing) is also interesting: the child did not look at the robot, and instead
remained focused on the secondary tablet. This situation is typically useful for
the robot to detect as it may want to adapt its behavior to recover the child's
attention.

@@

We have presented how a robot can
effectively assess in real-time with a regular camera, the focus of attention
of its interactant, and how we can combine it with the robot's {\it a priori}
knowledge
about the interaction to build a metric of
\emph{with-me-ness} over the course of the interaction.

@@

We already stated the main limitations of our approach to estimating the
focus of attention: eye gaze information is neglected and we do not perform
visibility check of the in-focus objects (we simply approximate them to
their origins, ignoring possible occlusions).

While the first issue is shared with most of the other vision (2D or 3D) or motion
capture techniques for real-time gaze estimation found in robotics,
our results are positive: we show that relying
purely on head pose estimation to estimate gaze direction leads to real-world
measures that are worth being considered and used. They may not match manual
annotations, but they are definitely a valuable \emph{in-the-moment} input for
the robot. For certain children, we reach levels of accuracy traditionally
considered as good.

@@

Our approach relies on a simple, non-intrusive sensor (a RGB camera by
the robot) and an open-source, fast pose estimation algorithm\fixme{test on nao!
how fast is it actually?}: we hope that this may contribute to the widespread
adoption of such a technique on a range of robots, including the relatively common {\sc
nao} platform.

@@

Borrowing the neologism from the field of CSCL, we have also introduced in this
article
\emph{with-me-ness} as a measure of ``how much the user is \emph{with} the robot during
a task''. This can be acquired over the course of the interaction,
thus providing the robot with a real-time metric for a relatively high-level social construct,
undoubtedly related to engagement.

One may reasonably wonder how different \emph{with-me-ness} is from \emph{joint
attention} on one hand, and from \emph{engagement} on the other hand.
With-me-ness is related to both, with however noteworthy nuances: (Triadic) \emph{joint
attention} is understood as the cognitive realization of a shared attention to
an object, itself building on a shared perception of that object (\ie joint attention
builds on a \emph{perceptual} alignment of two agents). \emph{Conceptual
with-me-ness} as proposed by Sharma~\etal in~\cite{sharma2014me} is on the contrary \emph{referential}: ``you are
\emph{with} me if you focus on what I refer to, either explicitly or
implicitly''. We understand it here in a slightly broader sense that
reflects the \emph{interaction}: ``you are \emph{with} me if you focus on
what is important for the interactive task at hand.''

On the other hand, with-me-ness is only a precursor of engagement: it does not
say much about the \emph{cognitive} commitment of a user to an interaction. A
user may closely adhere to the injunctions of the robot (or, actually, of the
experimenters), with thus high levels of with-me-ness,  without being
\emph{engaged} in the interaction. This is typically seen in child-robot
interaction: children will attempt to closely follow what they are asked to do
-- which may \emph{look like} they are engaged in the interaction -- while
they merely \emph{obey orders}.

Compared to engagement, one of the strengths of with-me-ness is its specificity:
it is well-defined (the only important variable -- that needs to be reported --
is the mapping {\it phase/expected attentional target}, \ie our
Table~\ref{tab:targets}), we can formalize it, and as such, it is valuable to
assess and compare how users are willing or able to interact with a robot.
We have hopefully demonstrated in this article that with-me-ness is an
operational \emph{in-the-moment} metric that can be also used as feedback to the
robot controller to build richer, more adaptive interactive behaviors for our
robots.
